---
title: 'Intro to Networks '
author: G
date: '2020-03-08'
slug: intro-to-networks
categories: []
tags: []
description: ''
---

```{r setup, include=FALSE, fig.align='center'}
#knitr::opts_chunk$set(echo = TRUE)
```

# Libraries needed:
```{r message=FALSE, warning=FALSE}
library(tidygraph) # Network manipulation
library(magrittr)  # Data wrangling
library(tidyverse) # Data wrangling
library(ggraph)    # Static network plotting
library(patchwork) # Arrainging satic plots
library(networkD3) # Interactive network plotting
```


# Lets make a simple graph!
Making networks is way easier than it sounds. We can start with two simple vectors of equal length and bind them into a dataframe of two columns.
```{r warning=FALSE}
from  <- c(1, 1, 2, 3, 4, 5)

to    <- c(2, 3, 4, 4 ,5, 6)

# combine the from and to vectors into a simple dataframe
basic <- tibble(from, to)
```
So what we have is an edge list that says "there is a connection from node 1 to node 2, anther from node one to node 3." And on and on... 
```{r echo=FALSE}
basic
```
Now lets use networkD3's `simpleNetwork()` function in order quickly make an interactive plot. The nice thing about this function is that it only requires an edge list. There is no wrangling needed. Go on, click and drag a node!
```{r warning=FALSE}
networkD3::simpleNetwork(basic)
```
This an awesome function, but the functionality and customization is a bit limited. We will use another function from this package to leverage more customization later on in the post. 

# Converting the edge data frame 
In order to convert the dataframe we use `tidygraph::as_tbl_graph()`. This will convert our simple edge list into a tidygraph object.
```{r messages = FALSE}
tidy_basic <- basic %>% 
  as_tbl_graph()
```

The tidygraph object is two dataframes connected. The nodes (each point) and edge (each connection).
```{r messages = FALSE, echo=FALSE}
tidy_basic
```


In order to manipulate the tbl_graph we need to tell R which of the two data frames we want to operate on. This is done with `activate(nodes)` or `activate(edges)`. Here we will name each node with a letter, and give each edge a random amount of transparency. 
```{r messages = FALSE}
tidy_basic_manipulated <- tidy_basic %>% 
  activate(nodes) %>%              # tell tidygraph we want to work on the modes dataframe
  mutate(id = letters[1:6]) %>%    # adding an id column for each node a - f
  activate(edges) %>%              # tell tidygraph we want to work on the edge dataframe
  mutate(alpha = runif(6, min = 0, max = 1)) # give each edge a random transparency value
```
Let's look at the columns we added to the plot.
```{r messages = FALSE, echo=FALSE}
tidy_basic_manipulated
```

Some quick plotting using ggraph. We will delve into this a bit more later. We need to specify the layout, and we need to give both the nodes and edges a `geom`.
```{r messages = FALSE}
ggraph(tidy_basic_manipulated, layout = 'sugiyama')+
  geom_edge_link  (aes(alpha = alpha)) +
  geom_node_point (aes(color = id)) + 
  geom_node_text  (aes(label = id), repel = TRUE)

```
Great we created a simple edge list, made it into a tidy graph object, manipulated it, then plotted it. Next we will step up the complexity but still 

# A slightly more complex example
Lets look a the highschool dataset from `ggraph`. Now this data has a yearly component, we'll get to that part but for now lets stick to one year for simplicity.  
```{r}
hs_tidy <- highschool %>% 
  filter(year == 1957) %>% # remove the time component for this example
  select(-year) %>% 
  as_tbl_graph()
```

And lets plot it real quick.
```{r}
ggraph(hs_tidy, layout = "stress") +
  geom_node_point() +
  geom_edge_link()
```



Now that we have the data into tidygraph's tbl_graph object we can easily calculate many statistical measures for each node edge easily combining `activate()` and `mutate()`. There are quite a few centrality measures a full list can be found [HERE](https://rdrr.io/cran/tidygraph/man/centrality.html). Thankfully tidygraph gives us nice clean wrappers for all the centrality measures created in `igraph`.
```{r}
hs_manipulated <- hs_tidy %>% 
  activate(nodes) %>% 
  mutate(  
    # calculate various centrality measures
    central_alpha  = centrality_alpha(), 
    central_degree = centrality_degree(),
    # calculate various grouping/clustering measures
    components     = as.factor(group_components()), 
    walktrap       = as.factor(group_walktrap()) 
  ) %>%    
  activate(edges) %>% 
  mutate(  # calculate edge centrality
    weight =  centrality_edge_betweenness()
  )
```

Let's take a look at our new metrics
```{r messages = FALSE, echo=FALSE}
hs_manipulated %>% activate(nodes)
```


Let's plot it and specify aesthetics based off the metrics we calculated.
```{r}
ggraph(hs_manipulated, layout = 'stress') +
  geom_edge_link (aes(alpha = weight)) +
  geom_node_point(aes(color = walktrap,
                      size = central_alpha)) 

```


Now lets play around a bit with some of the functionality of ggraph. This package is made to behave a lot like ggplot2. However, one key difference is the layout argument. These tell ggraph what algorithm to use when laying out the nodes. You can really go as deep as you want down this rabbit hole if you want to start playing around the with the parameters of the layout look at the [particles package](https://cran.r-project.org/web/packages/particles/vignettes/intro.html). Or, you can just do what I do and choose random ones until you get one you like.

```{r}
a <- ggraph(hs_manipulated, layout = "kk") +
  geom_node_point(aes(color = components)) +
  labs(title = "Layout stays the same",
       subtitle = "Nodes don't move")

b <- ggraph(hs_manipulated, layout = "kk") +
  geom_edge_link() +
  geom_node_point(aes(color = components)) 
  

c <- ggraph(hs_manipulated, layout = "kk") +
  geom_edge_hive() +
  geom_node_point(aes(color = components)) 


d <- ggraph(hs_manipulated, layout = "linear") +
  geom_edge_arc() +
  geom_node_point(aes(color = components)) +
  labs(title = "Change in layout",
       subtitle = "Move the nodes")

e <- ggraph(hs_manipulated, layout = "linear", circular = TRUE) +
  geom_edge_arc() +
  geom_node_point(aes(color = components)) 
  

f <- ggraph(hs_manipulated, layout = "mds") +
  geom_edge_link() +
  geom_node_point(aes(color = components)) 


# laying out the plots with patchwork
plots_compared <- ((a/b/c) | (d/e/f)) +
  plot_layout(guides = 'collect')


plots_compared
```



# Use networkD3 to make a more detailed plot

So, unfortunately NetworkD3 does not play nicely directly with tidygraph so we need to do some finagling. First we need to convert our tidygraph object to a network d3 object using `igraph_to_networkD3()`. In order to specify the group each nodes belongs in we need to extract a named vector to specify the `group =` argument. Then we need to create our JavaScript color scale. 
```{r warning=FALSE, messages = FALSE}
group_information <- hs_manipulated %>% 
  activate(nodes) %>% 
  as_tibble() 

node_vec <- as.numeric(group_information[["name"]])
group_vec <- as.numeric(group_information[["components"]])


group_info <- set_names(x = group_vec, nm = node_vec)

# Convert igraph to list for networkD3
d3_object <- igraph_to_networkD3(hs_manipulated, group = group_info)

# Set the colors for d3 orange and blue for now
ColourScale <- 'd3.scaleOrdinal()
            .domain(["1", "2"])
           .range(["#FF6900", "#89cff0"]);'



# Plot as a forceDirected Network
forceNetwork(Links = d3_object$links, 
             Nodes = d3_object$nodes, 
             Source = 'source',
             Target = 'target', 
             NodeID = 'name', 
             Group = 'group',
             colourScale = JS(ColourScale),
             zoom = TRUE)

```




